# The R-team attempts ANOVA and its non-parametric alternatives

## Achievements to unlock

Now that Leslie is getting more familiar with R coding, Nancy thinks it is time to learn one of the biggest strengths of R, making graphs! As the team saw in the previous chapter, graphs are very important for adding context to the numbers. For example, the negative skewness value for the PHYSHLTH variable in the previous chapter could have led an analyst to believe the variable was left skewed rather than right since negative skewness *almost always* means left skewed. A quick histogram plot showed otherwise and allowed the team to interpret the skewness accurately and identify the appropriate descriptive statistics to use for the situation. In addition to clarifying statistical results, graphs are one of the best way to communicate about data with various audiences. 

This chapter will primarily use the <span style="font-family:Lucida Console, monospace;font-weight:bold">ggplot2</span> package, which is part of the <span style="font-family:Lucida Console, monospace;font-weight:bold">tidyverse</span> to create and format common graphs used to display data. The chapter will cover which graphs are appropriate for different data types, the important features of a well-formatted graph, ways to avoid creating a misleading graph, and how to interpret graphs. Nancy and Kiara made a list of tasks for Leslie: 

T-test work great for comparing means in two groups, but what happens when there are three or more groups?  

* Achievement 1: Using graphics and descriptive statistics to make a prediction
* Achievement 2: Understanding and conducting one-way Analysis of Variance (ANOVA) 
* Achievement 3: Choosing and using post-hoc tests and contrasts
* Achievement 4: Testing ANOVA assumptions
* Achievement 5: Choosing and using alternative tests when assumptions are not met
* Achievement 6: Understanding and conducting two-way ANOVA

Follow Chelsea, Bobbi, and Leslie through the examples and exercises to test some relationships.

## The tech use problem

On her way to meet Chelsea and Leslie, Bobbi's cell phone battery is getting low. She turns off the phone and puts it in her backpack until she can get somewhere to charge it. While she continues her walk to campus, she notices that almost every person she walks past is using their cell phone. She starts thinking about her cell phone use and all the ways she is looking at screens and using technology throughout the day. Certainly it seems very different than her life just 10 years earlier. She thinks about the time she spends with her friends and how they all seem to be using phones or have phones nearby during gatherings. Her colleagues at work spend most of their time in front of computer screens in addition to having cell phones on. 

When she gets to her meeting with Chelsea and Leslie she asks them how much of their time awake they think they spend looking at a screen or using technology. Both of them say A LOT, well over half of the time, maybe closer to 70 or 80 percent. Nancy mentions that one of her favorite forms of exercise is walking to play Pokemon GO and Bobbi describes listening to podcasts whenever she has a spare moment and playing app games like Tetris and Candy Crush when in line at the grocery store or elsewhere. Leslie remembers being on vacation and the sheer panic and disconnectedness that she felt when she thought she had lost her cell phone only to find it 10 minutes later under the seat of the Uber she was in (that she had requested using her phone). This story reminds Nancy of when she dropped her phone while riding bikes on vacation and spent a panicked half-hour retracing every step only to be extremely relieved to find that it had been turned in at a restaurant by a good human.

After this discussion they start to look up technology use and how it may be influencing the culture and the health and mental health of people. Leslie starts reading about how text messaging has become common and tolerated in social situations while at the same time viewed as an interruption that can signal a lack of care by the texter for others who are present [@cahir2015people]. Bobbi finds some research linking cell phone use to college students having lower satisfaction with life [@lepp2014relationship], lower test scores and grade point average [@lepp2015relationship; @bjornsen2015relations], and more sedentary behavior [@barkley2016mobile]. A couple of studies found that cell phone use was associated with stress, sleep disorders, and depression symptoms among recent college graduates [@thomee2011mobile; @rosen2016sleeping]. 

Chelsea starts looking for some publicly available data on technology use and finds a technology use variable in the General Social Survey (GSS). She downloads a GSS data set with 2018 data including demographics and a variable measuring the percent of time the survey participant estimates using technology each day. Bobbi goes ahead and loads and summarizes the data. Because the data are saved in an R data file with the file extension .rda, loading the data only requires the `load()` command.  

```{r}
# load GSS rda file
load("data/gss2018.rda")
```

Leslie notices that the data frame is automatically named GSS. She thinks it might be good to specify the year the data were collected by changing the data frame name to gss.2018 and changes the name of a data frame by assigning the data frame to a new object and then removing the old object:

```{r}
# assign GSS to gss.2018
gss.2018 <- GSS

# remove GSS
rm(GSS)
```

Once the data are loaded and well-named, it is time for data management. 

### Data management

Leslie gets out the codebook to look at how the variables of interest were coded. Much of the research seemed to be with younger adults who were in college, which could be a unique group given their high level of education and the existence of cell phones for a large part of their lives. With this in mind, the three decide to examine five variables: USETECH, HAPPY, SEX, AGE, DEGREE. Leslie uses `summary(gss.2018)` to examine the variables. 

```{r}
# examine the variables
summary(gss.2018)
```

She sees that the USETECH variable has a minimum value of -1 and a maximum of 999. The question used to get this variable was: 

> During a typical week, about what percentage of your total time at work would you normally spend using different types of electronic technologies (such as computers, tablets, smart phones, cash registers, scanners, GPS devices, robotic devices, and so on)?

The valid responses should be between zero and 100 percent of the time, so Leslie decides to recode values outside this range to be `NA`. She examines the codebook and it shows that the two values outside the range are -1 and 999. Using the `mutate()` command with `na_if()` from tidyverse, she recodes these two values and checks her results:

```{r}
# recode USETECH to valid range
library(tidyverse)
gss.2018 <- gss.2018 %>% 
  mutate(USETECH = na_if(USETECH, -1)) %>%
  mutate(USETECH = na_if(USETECH, 999)) 

# check recoding
summary(gss.2018$USETECH)
```

Although it was not shown in the codebook, it appears there is also a value of 998 that is outside the valid range of values. Happy that she found it before it messed up her results, Leslie adds another line of code to her data management and tries again:

```{r}
# recode USETECH to valid range
gss.2018 <- gss.2018 %>% 
  mutate(USETECH = na_if(USETECH, -1)) %>%
  mutate(USETECH = na_if(USETECH, 999)) %>%
  mutate(USETECH = na_if(USETECH, 998)) 

# check recoding
summary(gss.2018$USETECH)
```

This looks much better. Leslie notes that there are a lot of `NA` values for this variable. The other variables of interest are: AGE, DEGREE, SEX, and HAPPY. Leslie looks at the codebook for age and finds: 

* 89 = 89 or older 
* 98 = "Don't know" 
* 99 = "No answer"

She decides to recode the 98 and 99 responses to be `NA`. Instead of starting new data management code, she just adds on to the existing code:

```{r}
# recode USETECH and AGE to valid ranges
gss.2018 <- gss.2018 %>% 
  mutate(USETECH = na_if(USETECH, -1)) %>%
  mutate(USETECH = na_if(USETECH, 999)) %>%
  mutate(USETECH = na_if(USETECH, 998)) %>% 
  mutate(AGE = na_if(AGE, 98)) %>% 
  mutate(AGE = na_if(AGE, 99))

# check recoding
summary(gss.2018)
```

The three other variables, SEX, DEGREE, and HAPPY are categorical variables. The codebook shows some categories that might be better coded as `NA`: 

* DEGREE 
   + 8 = "Don't know" 
   + 9 = "No answer" 
* HAPPY
   + 8 = "Don't know"
   + 9 = "No answer" 
   + 0 = "Not applicable"
   
```{r}
# recode variables of interest to valid ranges
gss.2018 <- gss.2018 %>% 
  mutate(USETECH = na_if(USETECH, -1)) %>%
  mutate(USETECH = na_if(USETECH, 999)) %>%
  mutate(USETECH = na_if(USETECH, 998)) %>% 
  mutate(AGE = na_if(AGE, 98)) %>% 
  mutate(AGE = na_if(AGE, 99)) %>% 
  mutate(DEGREE = na_if(DEGREE, 8)) %>% 
  mutate(DEGREE = na_if(DEGREE, 9)) %>% 
  mutate(HAPPY = na_if(HAPPY, 8)) %>% 
  mutate(HAPPY = na_if(HAPPY, 9)) %>% 
  mutate(HAPPY = na_if(HAPPY, 0))

# check recoding
summary(gss.2018)
```

The variables for analysis look like they all contain valid response options now, but every variable appears to be either numeric data type or integer data type. For DEGREE, HAPPY, and SEX, the data type should be factor. Leslie also prefers to have the category labels in her factors rather than just numbers. She reviews the codebook and makes these changes to the data:

```{r}
# recode variables of interest to valid ranges
gss.2018 <- gss.2018 %>% 
  mutate(USETECH = na_if(USETECH, -1)) %>%
  mutate(USETECH = na_if(USETECH, 999)) %>%
  mutate(USETECH = na_if(USETECH, 998)) %>% 
  mutate(AGE = na_if(AGE, 98)) %>% 
  mutate(AGE = na_if(AGE, 99)) %>% 
  mutate(HAPPY = na_if(HAPPY, 8)) %>% 
  mutate(HAPPY = na_if(HAPPY, 9)) %>% 
  mutate(HAPPY = na_if(HAPPY, 0)) %>% 
  mutate(SEX = factor(SEX, labels = c("male","female"))) %>%
  mutate(DEGREE = factor(DEGREE, labels = c("< high school",
                      "high school", "junior college",
                      "bachelor", "graduate"))) %>% 
  mutate(HAPPY = factor(HAPPY, labels = c("very happy", "pretty happy",
                                          "not too happy")))
# check recoding
summary(gss.2018)
```

## Achievement 1: Using graphics and descriptive statistics to make a prediction

Chelsea and Bobbi predict that people with higher educational degrees use technology more. They check this with the mean and standard deviation computed for each group. Instead of using `mean()` and `sd()` with `by()` or even `compmeans()`, they try using the tidyverse `group_by()` command:

```{r}
# mean and sd of age by group
use.stats <- gss.2018 %>% 
  group_by(DEGREE) %>%
  summarize(m.techuse = mean(USETECH, na.rm = T),
            sd.techuse = sd(USETECH, na.rm = T))
use.stats
```

Chelsea and Bobbi think they may be right about tech use increasing with higher degree groups. The mean portion of time using tech gets higher with each degree group. Leslie notices that the standard deviations are quite large and, for the less than high school group, the standard deviation is larger than the mean. She thinks she has read somewhere that this could be a problem for analyses that require normal distributions. 

```{r fig.cap="Distribution of tech use by degree."}
# graph usetech
ggplot(data = subset(gss.2018, !is.na(USETECH)), 
       aes(y = USETECH, x = DEGREE)) + 
  geom_boxplot(aes(fill = DEGREE), alpha = .8) + 
  geom_jitter(aes(color = DEGREE), alpha = .3) +
  scale_fill_manual(values = c("gray40", "dodgerblue2","#88398a", 
                               "deeppink", "gray80"), guide = FALSE) +
  scale_color_manual(values = c("gray40", "dodgerblue2","#88398a", 
                                "deeppink", "gray80"), guide = FALSE) +
  theme_minimal() + xlab("Degree")

  
```

Leslie notices that a lot of people in the first two category have selected zero percent of their time is on technology. She also notes that for all but the first category, there are a lot of people who have selected 100% of their time is spent using technology. Leslie asks Chelsea and Bobbi about these observations and how they might influence the results. Chelsea says there are terms for measures where people are at the very top or very bottom of the range. When there are a lot of observations at the very bottom of a range, this is called a **floor effect**, while seeing a lot of observations at the top of a range is called a **ceiling effect**. 

Leslie asks if they can still use ANOVA when there are floor and ceiling effects. Chelsea says yes, but with some caution. She explains that, when there are floor or ceiling effects, this means that the variation in a measure is limited by its range. Since ANOVA is an analysis of **variance** which examines central tendency and variation together, the limitations of floor and ceiling effects can result in not finding differences where there are differences.

Chelsea mentions that the limitation with ceiling and floor effects usually is a problem when the underlying measure has a wider range than what is measured. In the case of technology use, the range of zero to 100 percent of the time is as wide as it can be, so the observations at the ceiling and floor of this measure are just accurately reflecting very low and very high levels of technology use among many in the sample. 

### Unlock achievement 1: Check your understanding

Summarize the different parts of a boxplot including the upper and lower bounds of the box, the middle line through the box, and the whiskers. 

## Achievement 2: One-way analysis of variance (ANOVA)

The t-tests were great for comparing two means, but Leslie wonders what happens when there are three or more means. For example, examining mean technology use across degree groups. Would she just do a t-test for each pair of means and compare mean tech use for those with less than high school to high school, less than high school to junior collect, and so on? Chelsea explains that conducting multiple tests on the same two variables is not a great idea given that each statistical test come with some probability that the interpretation is incorrect. 

### Multiple t-tests and the familywise error rate

For example, with the first independent samples t-test from the prior chapter, the probability that the two group means were the same in the populations the samples came from was .0000000000002886. Although this is not a large probability, it is also not zero. If the threshold for the p-value to indicate statistical significance is .05, there can be up to a 5% probability that _rejecting the null hypothesis_ is an error. Rejecting the null hypothesis when it is true is called a _Type I error_ (see pull out box). 

With five groups in the degree variable, comparing each pair with a t-test (i.e., conducting _pairwise comparisons_) would result in 10 t-tests. If each t-test had a p-value threshold of .05 for statistical significance, the probability of at least one Type I error is fairly high. The formula for this probability of a Type I error when there are multiple comparisons is: 

$$
\begin{equation}
\alpha_{f}=1 - (1-\alpha_{i})^k
 (\#eq:cov)
\end{equation}
$$

Where $\alpha_{f}$ is the familywise Type I error rate, $\alpha_{i}$ is the individual alpha set as the statistical significance threshold, and c is the number of comparisons. The formula for computing c is k*(k-1)/2, where k is the total number of groups.

So, for a five group degree variable where we set the alpha for each pairwise comparison at .05, the familywise alpha would be:

$$
\begin{equation}
\alpha_{f}=1 - (1-.05_{i})^{10}=.40
 (\#eq:cov)
\end{equation}
$$

So, with 3 pairwise comparisons, the familywise $\alpha$ indicates there is a 40% probability that a conclusion was wrong on at least one of the t-tests. To control this error rate, and for efficiency, use a single analysis of variance (ANOVA) test instead of 10 t-tests. ANOVA is useful for testing whether three or more means are equal. It can be used with two means, but the t-test is preferable because it is more straightforward.

### The F test statistic for ANOVA

To compare mean age across the five degree categories, Chelsea introduces Leslie to one of the  ANOVA commands in R, `aov()`. The `aov()` command is in the **stats** library that loads with base R, so there are no new packages to install and load.

```{r}
# mean tech use percent by degree groups
techuse.by.deg <- oneway.test(formula = USETECH ~ DEGREE, 
            data = gss.2018, 
            var.equal = TRUE)
techuse.by.deg
```

The output shows a couple of familiar statistics and one new one, the _F_ statistic. The _F_ statistic is a ratio where the variation _between_ the groups is compared to the variation _within_ the groups. The _between group variation_ is in the numerator to calculate F, while the _within group variation_ is in the denominator. Leslie is a little confused by this, so Bobbi suggests modifying the boxplot to help think it through:

```{r fig.cap="Distribution of tech use by degree."}
# visualize variation between and within groups
ggplot(data = subset(gss.2018, !is.na(DEGREE)), 
       aes(y = USETECH, x = DEGREE)) + 
  geom_jitter(aes(color = DEGREE), alpha = .3) +
  scale_color_manual(values = c("gray40", "dodgerblue2","#88398a", 
                                "deeppink", "gray80"), guide = FALSE) + 
  stat_summary(fun.x = mean, color = "#88398a", geom = "tile", size = .65) + 
  geom_hline(yintercept = mean(gss.2018$USETECH, na.rm = TRUE), size = 1.25,
             color = "#fdb863") +
  theme_minimal() + xlab("Degree")
```

Leslie notices that the group mean for the junior college group (purple line) is higher than the overall mean (orange line). That is, the mean percent of time people use tech in this group is higher than the overall mean percent of time people use tech. The group mean does a better job than the overall mean of explaining the tech use distribution in each group. The difference between the group mean and the overall mean is _how much better_ the group mean is at representing the group. This difference is used to compute the numerator of the F statistic:

$$
\begin{equation}
F=\frac{\frac{\sum{n_{i.}(\overline{y}_{i.}-\overline{y})^2}}{k-1}}{\frac{\sum{\sum(y_{ik}-\overline{y}_{i.})^2}}{n-k}}
 (\#eq:cov)
\end{equation}
$$

Where _n_ represents an individual observation, _y_ stands for the outcome variable. The subscript _i_ is for individual and the subscript _k_ is for group. Breaking down the formula you can see that, in the numerator, there is $\overline{y}_{i.}-\overline{y}_{}$, with $\overline{y}_{i.}$ being the mean of each group and $\overline{y}$ being the overall or **grand** mean. Subtracting the overall mean from the group mean results in the difference between the group and the overall sample; this can be positive or negative and so is squared for the sum to more accurately represent the difference. This squared value is then multiplied by $n_{i.}$ or the number of people in the group and divided by $k-1$ or the number of groups minus 1. This results in a numerator that quantifies the overall difference between group means and the grand mean for all the participants in the sample.

The denominator sums the squared difference between each individual observation $y_{ik}$ and the mean of the group the individual is in, quantifying how far the individuals in the group are from the mean of the group. This is divided by the number of individuals minus the number of groups. 

Altogether, the numerator quantifies the variation between the group means and the grand mean, while the denominator quantifies the variation between the individual values and the group means. The statistic is sometimes simplified as:

$$
\begin{equation}
F=\frac{between-group-variability}{within-group-variability}
 (\#eq:cov)
\end{equation}
$$

Because the difference between the group means and the grand mean represents the variability that the group mean explains for the group, the numerator is also sometimes referred to as **explained variance**. With the denominator summing the remaining distances, or information that is not explained by the group mean, between the observations and their group mean, the denominator is sometimes called **unexplained variance**. The F statistic, then, can be referred to as a ratio of explained to unexplained variance. That is, how much of the variability in your data does the ANOVA model explain compared to how much it leaves unexplained. So, the larger the F-statistic, the more the ANOVA has explained compared to what it has left unexplained.

Leslie notices that there are clearly some observations closer to the overall mean than they are to the group mean for this group and for other groups. Bobbi explains that this is true for some observations, but the differences between each observed value and the group mean added together is a smaller number than the differences between each observed value and the overall mean. 

True to the _analysis of variance_ name, the F-statistic can also be represented as the ratio of the variance between the groups to the variance within the groups:

$$
\begin{equation}
F=\frac{s_b^2}{s_w^2}
 (\#eq:cov)
\end{equation}
$$

Once the F-statistic is computed, the probability of finding an F at least as big as the one computed is determined by the F-distribution:

```{r echo = FALSE, fig.cap="F distribution (df = 4 and 1404) with shaded rejection region for .05."}
x <- seq(0,60,length=10000)
db <- df(x, 4, 1404)
dat <- data.frame(x, db)
qplot(x, db, data = dat, geom = "line") +
  geom_line()+
  theme_minimal() +
  geom_area(mapping = aes(x = ifelse(x > techuse.by.deg$statistic, x, 0)), fill = "#88398a") +
  ylim(0,1.1) + xlim(0, 45) +
  xlab("F statistic") + ylab("Probability") 
```

Chelsea mentions that the F distribution may remind Leslie of the chi-squared distribution from the previous chapter since they are both right-skewed and do not go below zero. Also like the chi-squared distribution, the F distribution shape changes based on the degrees of freedom. 

The purple shaded area under the curve represents the top 5% of F-statistic values. The ANOVA for had an F-statistic of `r techuse.by.deg$statistic`, so it is far to the right in the distribution. The probability of an F-statistic this large or larger is `r format(techuse.by.deg$p.value, scientific = FALSE)`. With a p-value this tiny, the result would be considered statistically significant. Bobbi suggests going through the steps of NHIST to organize the analyses more clearly. 

#### NHST for ANOVA

##### NHST Step 1: Write the null and alternate hypotheses

HO: The mean time spent on technology use is equal across degree groups.

HA: The mean time spent on technology is not equal across these groups.

##### NHST Step 2: Compute the test statistic

```{r}
# print the results of the ANOVA
techuse.by.deg
```

##### NHST Step 3: Compute the probability for the test statistic (p-value) 

The p-value in this case is `r format(techuse.by.deg$p.value, scientific = FALSE)`. The value of an F-statistic being this large or larger happens about `r 100*as.numeric(format(techuse.by.deg$p.value, scientific = FALSE))`% of the time when the null hypothesis is true. 

##### NHST Steps 4 & 5: Interpret the probability and write a conclusion 

The mean time spent on technology use was significantly different across degree groups [F(`r techuse.by.deg$parameter[1]`,`r techuse.by.deg$parameter[2]`) = `r round(techuse.by.deg$statistic)`; p = `r format(techuse.by.deg$p.value, scientific = FALSE)`] indicating these groups likely came from populations with different mean time spent on technology use. The highest mean was `r use.stats[5,1]`% of time used for technology for those with graduate degrees. The lowest mean was `r use.stats[1,1]`% of the time for those with less than a high school diploma. 

### Unlocking achievement 2: Check your understanding

Analysis of variance, or ANOVA, is useful for:

* Comparing the means for three or more groups  
* Comparing a single mean to a population or hypothesized value 
* Transforming a categorical variable into a continuous variable  
* Examining the relationship between two categorical variables 

## Achievement 3: Post-hoc tests and planned contrasts

Leslie asks Bobbi if they can report that the **graduate degree** group has a significantly higher mean time spent using technology than people in the other groups. Bobbi explains that ANOVA is an _omnibus_ test, which means it identifies whether there are any differences, but doesn't give any information about what is driving the significant results. Chelsea reminds Leslie that another example of an _omnibus test_ is chi-squared. While they used chi-squared to determine whether a relationship was statistically significant, they had to compute standardized residuals or an odds ratio to make any claims about what the relationship is. Chelsea mentions that there are two main ways to determine where significant differences among groups are following a significant omnibus test:

* Post-hoc tests 
* Planned contrasts 

The two methods are both useful for examining differences among means. The difference between post-hoc tests and planned contrasts is that post-hoc tests examine **each pair of means** to determine which ones are the most different from each other and are therefore driving the statistically significant results, while planned contrasts can compare subsets of means or groups of means. Leslie is confused by the planned contrasts idea. Bobbi thinks going through a couple examples of post-hoc and planned contrasts would be best to clarify.

### Post-hoc tests

#### Bonferroni 

There are several different types of post-hoc tests, one of the most commonly used is the _Bonferroni_ post-hoc test. The Bonferroni pairwise post-hoc test does a t-test for each pair of means but adjusts the threshold for statistical significance to ensure that there is a small enough risk of Type I error. 

Bobbi writes a line of code to conduct the Bonferroni post-hoc tests with the continuous variable as x, the marital status variable as g (for groups) and the p-value adjustment specified to be bonf for Bonferroni:

```{r}
# find differences in mean tech use by degree groups
bonf.tech.by.deg <- pairwise.t.test(x = gss.2018$USETECH, 
                g = gss.2018$DEGREE, 
                p.adj = "bonf")
bonf.tech.by.deg
```

The output is different from previous statistical testing. Instead of a test statistic like _t_ or _F_, the output is a matrix of p-values. While the calculation of the t-statistic for each pair of groups is the same as other independent samples t-tests, the corresponding p-value is _adjusted_ to correct for the multiple statistical tests that could lead to an inflated Type I error. 

Specifically, the Bonferroni adjustment multiplies each p-value from each t-test by the overall number of t-tests conducted. Leslie remembers that there were 10 pairwise comparisons, so these p-values have been multiplied by 10. Leslie notices that a few of the resulting p-values are 1.00000. Bobbi explains that the p-value cannot be over 1, so for p-values that are over 1 when adjusted by the multiplication, they are rounded to exactly 1. 

The adjusted p-values for seven of the t-tests fall below .05 and so indicate that the difference in mean time using tech between two groups is _statistically significant_. It looks like there are significant differences in mean time between _< high school_ and all of the other groups (p < .05); likewise there are significant differences in mean time using tech between _high school_ and all other groups. There are no differences among the means of the three college groups. Bobbi suggests reporting the differences might be more meaningful with the actual mean values included. She reprints the `use.stats` object:

```{r}
# mean age by groups
use.stats
```

Leslie summarizes the statistically significant differences identified in the Bonferroni post-hoc test:

Mean percent of time using technology is statistically significantly (p < .05) _lower_ for people with less than a high school diploma (m = `r round(use.stats[1,2], 1)`) compared to each of the other groups where the mean percent of time using technology ranges from `r round(use.stats[2,2], 1)` to `r round(use.stats[5,2], 1)`. 

#### Tukey's Honestly Significant Difference

Chelsea advises Leslie that the Bonferroni post-hoc test is generally considered a very conservative post-hoc test that only identifies the very largest differences between means as statistically significant. This might be useful sometimes, but not always. When a less strict test will work, Chelsea recommends using _Tukey's Test_. The Tukey post-hoc test is a modified t-test with the test statistic, _q_, calculated:

$$
\begin{equation}
q=\frac{m_1 - m_2}{SE}
 (\#eq:cov)
\end{equation}
$$

Leslie is perplexed since this appears to be the same exact formula as the t-statistic for the independent samples t-test. Bobbi explains that the q test statistic is the same as the t, but the q distribution is different from the t-distribution, raising the critical value necessary for the q statistic to reach statistic significance. So, even with the same test statistic, it is more difficult to reach statistical significance with a Tukey Test compared to a t-test.

Unfortunately, Chelsea explains, the `TukeyHSD()` command does not work well with the `oneway.test()` output, so the entire ANOVA model has to be re-estimated to use with the command. The `aov()` command works and takes similar arguments to the `oneway.test()` command, so nesting the estimation using `aov()` is one way to go:

```{r}
# Tukey post-hoc test for age.by.fake
tukey.tech.by.deg <- TukeyHSD(aov(formula = USETECH ~ DEGREE, 
                                  data = gss.2018))
tukey.tech.by.deg

```

Comparing these results to the Bonferroni test results above, the means that were statistically significantly different with Bonferroni are also statistically significantly different with the Tukey Test. However, the p-values are lower with Tukey. For example, the p-value for the comparison of bachelor to junior college was `r round(bonf.tech.by.deg[[3]][3,3], 1)` with the Bonferroni post-hoc test but `r round(tukey.tech.by.deg$DEGREE[8,4], 2)` with the Tukey post-hoc test. 

### Planned comparisons 

Rather than comparing every group to every other group, it might be more interesting to compare all the college groups as a whole to the other groups. Bonferroni and Tukey are not designed to do this, but there are other options. Planned comparisons can be used to compare one mean to another mean, two means to one mean, or really any subset of means to any other subset of means. 

Planned comparisons are computed by developing **contrasts** that specify which means to compare to which other means. For example, to compare all the college groups to high school group, the contrast would omit the less than high school group and compare the mean for everyone in the high school group to the mean of the combined three college groups.

```{r}
# look at the levels of education variable
levels(gss.2018$DEGREE)
```

The order of the factor variable is the order the contrast should be developed in. There are a few rules to keep in mind for a factor. First, the categories grouped together should all be represented by the same number in the contrast. Second, the numbers in the contrast should all add to zero. Finally, any category not included in the contrast should be represented by a zero. In this case, to compare the second level of the factor, **high school**, with the third, fourth, and fifth levels combined could be written as:

* 0 (do not include) 
* 3 (high school) 
* -1 (junior college) 
* -1 (bachelor) 
* -1 (graduate) 

The three categories represented by -1 will be grouped together. Adding 3 + -1 + -1 + -1 is equal to zero. 

```{r}
# put the contrast in a vector
contrast1 <- c(0, 3, -1, -1, -1) 
```

Bobbi suggests that Leslie look at the DEGREE variable in the environment window. The variable now has an attribute called **contrasts** that lists the zeros, threes, and negative ones. 

```{r}
# link the contrast to the categorical variable using contrast()
contrasts(gss.2018$DEGREE) <- contrast1

# re-run the model using aov() 
techuse.by.deg.aov <- aov(formula = USETECH ~ DEGREE, 
            data = gss.2018)

# apply the contrasts to the anova object techuse.by.deg.aov
# give the contrast a good descriptive name
tech.by.deg.contr <- summary.aov(techuse.by.deg.aov, 
            split = list(DEGREE = list("high school vs. college" = 1)))
tech.by.deg.contr

```
The output shows that mean technology use for those who finished high school is statistically significantly different from mean technology use for the three college groups combined [F(`r tech.by.deg.contr[[1]][2,1]`, `r tech.by.deg.contr[[1]][2,2]`) = `r tech.by.deg.contr[[1]][2,4]`; p < .001]. Leslie is interested in looking at the means being compared with this contrast. 

To create the graph Leslie uses the tidyverse to `mutate()` the DEGREE variable so it groups the three college groups into a single group. The less than high school group can be recoded to `NA` or something else since it will not be included in the graph or table. The unneeded values can be filtered out before making the table or plot using the `filter()` command.   

```{r}
# recode and compute the means for high school and college groups
gss.2018 %>%
  mutate(DEGREE = factor(DEGREE, labels = c(NA,
                      "high school", "college",
                      "college", "college"))) %>% 
  group_by(DEGREE) %>% 
    summarise(m.techuse = mean(USETECH, na.rm = T),
            sd.techuse = sd(USETECH, na.rm = T))
```

The difference is pretty large, a visual might help to add more context: 

```{r fig.cap="Distribution of tech use by degree for contrast."}
# recode and filter the variable so only the two degree categories
# are in the plot
gss.2018 %>% 
  mutate(DEGREE = factor(DEGREE, labels = c(NA,
                      "high school", "college",
                      "college", "college"))) %>% 
  filter(DEGREE == "high school" | DEGREE == "college") %>% 
  ggplot(aes(y = USETECH, x = DEGREE)) +
    geom_boxplot(aes(fill = DEGREE), alpha = .8) + 
    geom_jitter(aes(color = DEGREE), alpha = .3) +
    scale_fill_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    scale_color_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    theme_minimal() + xlab("Degree")

```

It seems clear to Leslie that these means are different from each other. She thinks the less than high school group is probably also different from the college groups. Chelsea says they can test both those ideas together using planned comparisons, like this:

```{r}
# less than HS v. all college contrast
contrast2 <- c(3, 0, -1, -1, -1)

# bind the two contrasts together into a matrix
cons <- cbind(contrast1, contrast2)

# connect the matrix with the factor variable
contrasts(gss.2018$DEGREE) <- cons

# estimate the ANOVA with contrasts
tech.by.deg.contr <- summary.aov(techuse.by.deg.aov, 
            split = list(DEGREE = list("high school vs. college" = 1, 
                                       "< high school vs. college" = 2)))
tech.by.deg.contr
```

Leslie graphs the data again again to better understand both contrasts. There is a clear visual difference between the high school group and the college group and also between the less than high school group and the college group:

```{r fig.cap="Distribution of tech use by degree for contrast comparing college to each of the other groups."}
# recode to group the college groups
gss.2018 %>% 
  mutate(DEGREE = factor(DEGREE, labels = c("< high school",
                      "high school", "college",
                      "college", "college"))) %>% 
  ggplot(aes(y = USETECH, x = DEGREE)) +
    geom_boxplot(aes(fill = DEGREE), alpha = .8) + 
    geom_jitter(aes(color = DEGREE), alpha = .3) +
    scale_fill_manual(values = c("dodgerblue2","#88398a","deeppink"), guide = FALSE) +
    scale_color_manual(values = c("dodgerblue2","#88398a","deeppink"), guide = FALSE) +
    theme_minimal() + xlab("Degree")

```

Leslie is interested in how many contrasts can be done and also how all these statistical comparisons might be inflating the Type I error that they discussed with the post-hoc tests. Leslie explains that, in addition to each comparison comparing just two things and each comparison adding to zero, the planned comparisons can, and should, isolate each group (e.g., the high school group) one time. This ensures that the contrasts are **independent** of each other since the variance for each group is only isolated and used in a statistical comparison one time. Because each group is isolated one time, the total maximum number of contrasts allowable is one less than the number of groups. Leslie wants to make sure she has this right and summarizes the important characteristics of contrasts:

* Contrast values add to zero 
* Each contrast compares two groups 
* Each group should be isolated one time 
* The maximum number of contrasts is one less than the number of groups 

Bobbi confirms that these are the characteristics of the contrasts. She explains that two things ensure you are following all the rules:

(1) Add up each contrast to make sure it adds to zero 

(2) Multiply each value for in each contrast with the corresponding values in the other contrasts and add up the products; this should also add to zero 

If each contrast adds to zero and the sum of the products across contrasts adds to zero, then the set of contrasts follows the rules for independence and can be used to understand differences.

Leslie notices that they have not yet developed a full set of contrasts that fit these criteria. So far they only have two contrasts. With five total degree groups, there should be four contrasts in the list. Leslie thinks about the interesting comparisons she could do and comes up with an idea for four comparisons:

* < high school v. high school and junior
* high school v. all three college groups 
* junior college v. bachelor's and graduate degrees  
* bachelor's v. graduate degree 

She creates the set of contrasts that correspond to these ideas:

```{r}
c1 <- c(2, -1, -1,  0,  0)
c2 <- c(0,  3, -1, -1, -1)
c3 <- c(0,  0,  2, -1, -1)
c4 <- c(0,  0,  0,  -1, 1)

conts <- cbind(c1, c2, c3, c4)
conts
```

Leslie looks at the matrix. She starts to check the values. The first column is the first contrast. When she adds up 2 + -1 + -1 + 0 + 0 it equals zero. She adds up the other three columns and they are all equal to zero. The first requirement is met. The second requirement is more complicated. She has to multiply each value across the four contrasts. The first value for the first contrast is 2, the first value of the second contrast is 0, the first value of the third contrast is 0, and the first value of the fourth contrast is also 0. Multiplying $2 * 0 * 0 * 0 = 0$. The product of the second values across the four contrasts is $-1 * 3 * 0 * 0 = 0$. The product of the third values is $-1*-1*2*0=0$, the fourth values multiply to $0*-1*-1*-1=0$, and finally the product across the fifth values in the contrasts is $0*-1*-1*1=0$. Adding all these products together results in $0*0*0*0*0=0$. So, the second requirement is met. The set of contrasts is independent and ready to use:

```{r}
# connect the matrix with the factor variable
contrasts(gss.2018$DEGREE) <- cons

# estimate the ANOVA with 4 independent contrasts
tech.by.deg.contr <- summary.aov(techuse.by.deg.aov, 
            split = list(DEGREE = list("< high school vs. high school & jr college" = 1, 
                                       "high school vs. college" = 2,
                                       "jr college vs. more college" = 3,
                                       "bachelor's vs. grad degree" = 4)))
tech.by.deg.contr
```

Leslie is excited that it worked! The code felt a little confusing so it is nice to see the result. She is wondering now whether the p-values should be adjusted to account for the multiple testing. Bobbi has done some reading on this and there does not seem to be consistent guidance, but since multiple statistical tests inflate the probability of a Type I error, she decides it is a good idea to apply some sort of correction. While this is not an option available in the `aov()` or `summary.aov()` commands, there is a `p.adjust()` command that adjusts p-values using one of several types of adjustments. Since Bonferroni is familiar and used in the post-hoc pairwise comparisons, Bobbi thinks it is a good place to start.

The first argument in the `p.adjust()` command is a p-value or a vector of p-values. The second argument is `method =`, which is where to specify "bonferroni." Bobbi gives it a try: 

```{r}
# adjust p-values for multiple comparisons 
adj.p.vals <- p.adjust(tech.by.deg.contr[[1]]$`Pr(>F)`, method = "bonferroni")
adj.p.vals

```

The adjusted p-values are still very small, so the conclusions about statistical significance have not changed, even when using a conservative adjustment like Bonferroni. 

Leslie wants one last visualization showing all the different comparison groups. It takes her a few minutes, but she ends up with a nice set of graphs demonstrating the differences among the various groups. She decides to try a violin plot instead of a boxplot this time. Violin plots the probability density of the data along the range of values and so might more clearly show some of the patterns in the data. Bobbi suggests that the difference between the violin plot and the boxplot is similar to the difference between a density plot and a histogram. The violin plot shows more subtle variations in the distribution of observations.

```{r echo = FALSE, fig.cap = "Visualizing the distribution of use by degree for four planned comparisons."}

c1graph <- gss.2018 %>% 
  mutate(DEGREE = factor(DEGREE, labels = c("< high school",
                      "high school & jr coll", "high school & jr coll",
                      NA, NA))) %>% 
  filter(DEGREE == "< high school" | DEGREE == "high school & jr coll") %>%
  ggplot(aes(y = USETECH, x = DEGREE)) +
    geom_violin(aes(fill = DEGREE), alpha = .8) + 
    geom_jitter(aes(color = DEGREE), alpha = .3) +
    scale_fill_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    scale_color_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    theme_minimal() + xlab("") + ylab("Percent of time using tech")

c2graph <- gss.2018 %>% 
  mutate(DEGREE = factor(DEGREE, labels = c(NA,
                      "high school", "college",
                      "college", "college"))) %>% 
  filter(DEGREE == "high school" | DEGREE == "college") %>%
  ggplot(aes(y = USETECH, x = DEGREE)) +
    geom_violin(aes(fill = DEGREE), alpha = .8) + 
    geom_jitter(aes(color = DEGREE), alpha = .3) +
    scale_fill_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    scale_color_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    theme_minimal() + xlab("") +ylab("")

c3graph <- gss.2018 %>% 
  mutate(DEGREE = factor(DEGREE, labels = c(NA,
                      NA, "jr college",
                      "more college", "more college"))) %>% 
  filter(DEGREE == "jr college" | DEGREE == "more college") %>%
  ggplot(aes(y = USETECH, x = DEGREE)) +
    geom_violin(aes(fill = DEGREE), alpha = .8) + 
    geom_jitter(aes(color = DEGREE), alpha = .3) +
    scale_fill_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    scale_color_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    theme_minimal() + xlab("Degree")+ ylab("Percent of time using tech")

c4graph <- gss.2018 %>% 
  mutate(DEGREE = factor(DEGREE, labels = c(NA,
                      NA, NA,
                      "bachelor", "master"))) %>% 
  filter(DEGREE == "bachelor" | DEGREE == "master") %>%
  ggplot(aes(y = USETECH, x = DEGREE)) +
    geom_violin(aes(fill = DEGREE), alpha = .8) + 
    geom_jitter(aes(color = DEGREE), alpha = .3) +
    scale_fill_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    scale_color_manual(values = c("dodgerblue2","#88398a"), guide = FALSE) +
    theme_minimal() + xlab("Degree") + ylab("")

# plots together
gridExtra::grid.arrange(c1graph, c2graph, c3graph, c4graph)

```

Leslie is still not convinced the bachelor's and master's degree categories should be statistically significantly different, but the violin helps to show the difference in the distributions a little more. Leslie asks Bobbi when it is most appropriate to use the post-hoc tests that compare each pair of means and when it is more appropriate to do planned comparisons. Bobbi explains that, when you have hypotheses ahead of time about which groups are different from one another, use planned comparisons. When you do not have hypotheses ahead of time about which means are different from each other, use post-hoc tests. 

### Unlocking achievement 3: Check your understanding

Why is a post-hoc or planned comparison test useful after a statistically significant ANOVA result? 

* There is no other way to know what the means are for each group   
* They confirm whether or not your ANOVA is statistically significant  
* Post-hoc and planned comparison tests help identify which groups means are different from each other   
* Post-hoc and planned comparison  tests ensure that both variables have minimal missing data for ANOVA  

Create another set of planned comparisons that follows the rules.

## Achievement 4: ANOVA assumptions

The first assumption for ANOVA is that the data are normally distributed. There are several ways to assess normality. Visually, a histogram or a QQ-plot is useful for identifying normal and non-normal data distribution. Statistically, a Shapiro-Wilk test can be used. 

### Testing normality 

Start with histograms:

```{r fig.cap="Distribution of technology use by degree categories."}
#graph tech use by degree 
ggplot(data = subset(gss.2018, !is.na(USETECH)), aes(x = USETECH)) + 
  geom_histogram(fill = "#88398a", col = "grey") + 
  facet_wrap(~ DEGREE, nrow = 2, scales = "free_y") + 
  theme_minimal() + 
  labs(x="Degree",
       y="Survey participants") 
  
```

The groups do not look normally distributed. The qq-plot might confirm:

```{r fig.cap="Distribution of technology use by degree."}
#graph systolic bp

ggplot(data = subset(gss.2018, !is.na(USETECH)), aes(sample = USETECH)) + 
  stat_qq(color = "#88398a") + 
  facet_wrap(~ DEGREE, nrow = 2, scales = "free_y") + 
  geom_abline(intercept = mean(gss.2018$USETECH, na.rm = TRUE), 
              slope = sd(gss.2018$USETECH, na.rm = TRUE)) +
  theme_minimal() + 
  labs(x="Percent time using tech",
       y="Survey participants") 

```

Leslie notes that none of the groups appear to be normally distributed based on the qq-plot. She also notices that the floor and ceiling values appear to be driving some of the non-normality. The Shapiro-Wilk test does not seem necessary in this case given the big deviations from normal in the histograms and qq-plots, however, Chelsea suggests that they try it just to confirm:

```{r}
# statistical test of normality for groups
by(gss.2018$USETECH, gss.2018$DEGREE, FUN = shapiro.test)
```

All five of the Shapiro-Wilk tests are statistically significant, indicating that the null hypothesis for this test (i.e., the data are normally distributed) is rejected in each group. 

### Homogeneity of variances assumption

The second assumption for ANOVA is the assumption of _homogeneity of variances_ or _equal variances across groups_. So, not only do the data need to be normally distributed, but the data should be equally spread out in each group. Leslie reviews the graphs thinks this might actually be an easier assumption to meet.

Chelsea mentions that the _Levene's Test_ is widely used to test the assumption of equal variances. The null hypothesis is that _the variances are equal_ while the alternate is that at least two of the variances are different. The **car** package in R has a Levene's test command that can be used to test for equal variances:

```{r}
library(car)

# equal variances for systolic by sex
leveneTest(y = USETECH ~ DEGREE, data = gss.2018)

```

The p-value for the Levene's test suggests rejecting the null hypothesis; the variances of USETECH are statistically significantly different across groups (p < .05). So, the ANOVA fails the assumption of _homogeneity of variances_.

Chelsea summarizes the assumptions for the tests to make sure Leslie has them all:

### ANOVA assumptions recap 

* continuous variable and three or more independent groups 
* independent observations 
* normal distribution in each group 
* equal variances for each group 

Leslie asks Chelsea what the difference is between independent observations and independent groups. Chelsea explains that independent observations is the assumption that the people in your data are not related to one another in any important way. Things that might violate this assumption are having siblings or spouses in a data set or measuring the same person multiple times. Independent groups is the assumption that two groups are not related to one another. If one group was made up of all the spouses of another group, the two groups would not be independent.

Leslie expresses some disappointment that ANOVA failed its assumptions. Bobbi explains that, like with the t-tests, there is an alternate statistical test to use when ANOVA assumptions are not met.

### Unlocking achievement 4: Check your understanding

Which ANOVA assumption(s) appear(s) to be violated by the data in the graph shown? 

```{r echo = FALSE, fig.cap="Distribution of data violating an ANOVA assumption."}
ggplot(data = subset(gss.2018, !is.na(HAPPY)), aes(x = USETECH)) + 
  geom_histogram(fill = "#88398a", col = "grey") + 
  facet_grid(HAPPY ~ .) +
  theme_minimal() 
```

* continuous variable and three or more independent groups 
* independent observations 
* normal distribution in each group 
* equal variances for each group 

## Achievement 5: Alternatives when ANOVA assumptions are not met

The two main assumptions for ANOVA have different alternatives 

### Calculating an alternate F-statistic for failing the homogeneity assumption

If the normality assumption is met but the homogeneity of variances assumption fails, the standard approach is to use ANOVA but compute an alternate F statistic that does not rely on equal variances. There are two alternate F-statistics that are widely used for this purpose:

* Brown-Forsythe 
* Welch 

#### Brown-Forsythe test 

The Brown-Forsythe approach to calculating F starts with a transformation of the continuous variable from its measured values to values that represent the distance each observation is from the median of the variable. So, had the homogeneity of variances assumption failed for comparing age across fake news perception groups, the age variable might be transformed this way:

$$
\begin{equation}
t = |y_{ij} - med_{y_{ij}}|
 (\#eq:cov)
\end{equation}
$$

The F-statistic is then computed using the same F statistic but with the means computed for the transformation (t) rather than the raw value (y) of the continous variable:

$$
\begin{equation}
F_{BF}=\frac{\frac{\sum{n_{i.}(\overline{t}_{i.}-\overline{t})^2}}{k-1}}{\frac{\sum{\sum(t_{ik}-\overline{t}_{i.})^2}}{n-k}}
 (\#eq:cov)
\end{equation}
$$

While there are R packages that can be used to compute the Brown-Forsythe directly, another option is to transform the outcome variable and use the `aov()` command used for ANOVA.

```{r}
# transform usetech variable
gss.2018$usetech.t <- abs(gss.2018$USETECH - median(gss.2018$USETECH, na.rm = T))

# brown-forsythe anova
usetech.t.by.degree <- oneway.test(formula = usetech.t ~ DEGREE, 
            data = gss.2018)
usetech.t.by.degree

```

The results show a statistically significant difference in the mean of the transformed usetech variable by degree group [F(`r usetech.t.by.degree$parameter[1]`, `r usetech.t.by.degree$parameter[2]`) = `r round(usetech.t.by.degree$statistic, 2)`; p = `r format(usetech.t.by.degree$p.value, scientific = FALSE)`]. Leslie is interested in how the transformed data used in these calculations looks; she decides to compute the descriptive statistics and examine a graph of the transformed variable to better understand the results.

```{r}
# means of transformed variable
usetech.t.stats <- gss.2018 %>% 
  group_by(DEGREE) %>%
  summarise(m.usetech.t = mean(usetech.t, na.rm = T),
            sd.usetech.t = sd(usetech.t, na.rm = T))
usetech.t.stats
```

The mean of the transformed usetech variable, which were the difference between participant usetech value and the median value of usetech, was `r round(usetech.t.stats$m.usetech.t[1], 2)` for the **< high school** group and `r round(usetech.t.stats$m.usetech.t[2], 2)` for the high school group. The rest of the means were smaller. The transformation made the differences somewhat smaller and the means higher in the lower education groups. A boxplot shows the differences across the groups:

```{r}
# graph age
ggplot(data = subset(gss.2018, !is.na(usetech.t)), 
       aes(y = usetech.t, x = DEGREE)) + 
  geom_boxplot(aes(fill = DEGREE), alpha = .8) + 
  geom_jitter(aes(color = DEGREE), alpha = .3) +
  scale_fill_manual(values = c("gray40", "dodgerblue2","#88398a", "deeppink", "gray80"), guide = FALSE) +
  scale_color_manual(values = c("gray40", "dodgerblue2","#88398a", "deeppink", "gray80"), guide = FALSE) +
  theme_minimal() + xlab("Degree") +
  ylab("Distance to median of tech use time")
```

Leslie notices that the transformation has reduced the differences between the junior college, bachelor, and graduate categories, with the boxplot showing nearly the same median value for the transformed variable. Chelsea reminds that the Welch's variation on ANOVA is also an option. 

While Leslie has been working on the Brown-Forsyth test, Bobbi looked up some additional information on the Welch test. Rather than use a transformed outcome variable, the main idea behind the Welch F-test is to use weights in the calculation of the group means and grand means. The weight is computed for each group, since violating the homogeneity of variances is due to different variances across groups:

$$
\begin{equation}
w_1 = \frac{n_k}{s^2_k}
 (\#eq:cov)
\end{equation}
$$

Where $n_k$ is the sample size in group k and $s^2_k$ is the variance in group k. The grand mean is computed using the weight and the weighted mean for each of the groups:

$$
\begin{equation}
\overline{y}_{welch} = \frac{\sum{w_k \overline{y}_{k}}}{\sum{w_k}}
 (\#eq:cov)
\end{equation}
$$

The Welch grand mean is then used in the final computation of the Welch F-statistic, which gets a little complex:

$$
\begin{equation}
F_{welch}=\frac{\frac{\sum{w_k(\overline{y}_k-\overline{y}_{welch})^2}}{r-1}}{1+\frac{2}{3}\Bigg(\frac{\frac{3\sum\bigg(1-\frac{w_k}{\sum{w_k}}\bigg)^2}{n_k-1}}{k^2-1}\Bigg)(k-2)}
 (\#eq:cov)
\end{equation}
$$

Leslie takes one look at the formula and asks if there is an R command to compute the value. Chelsea explains that the `oneway.test()` command computes Welch's F when the option `var.equal =` is set to be false, like `var.equal = FALSE`. Leslie gives it a try:

```{r}
# welch test for unequal variances
welch.usetech.by.degree <- oneway.test(formula = USETECH ~ DEGREE,
            data = gss.2018,
            var.equal = FALSE)
welch.usetech.by.degree
```

The results show another statistically significant difference in the mean of the usetech variable by degree group [F(`r welch.usetech.by.degree$parameter[1]`, `r welch.usetech.by.degree$parameter[2]`) = `r round(welch.usetech.by.degree$statistic, 2)`; p = `r format(welch.usetech.by.degree$p.value, scientific = FALSE)`]. Leslie notices that the F-statistic is a little larger and the degrees of freedom for the denominator is a smaller number. Chelsea explains that the weighting is also used in the calculation of the denominator degrees of freedom:

$$
\begin{equation}
df_{denom}=\frac{1}{\frac{3\sum\bigg(1-\frac{w_k}{\sum{w_k}}\bigg)^2}{n_k-1}}
 (\#eq:cov)
\end{equation}
$$

With fewer degrees of freedom, the F-statistic has to be a larger number to reach statistical significance. Examine a few F-distributions to see how this works:

```{r echo = FALSE, fig.cap="F-distributions with 4 degrees of freedom in the numerator."}
x <- seq(0,20,length=10000)
db2 <- df(x, 4, 2000)
db3 <- df(x, 4, 25)
db4 <- df(x, 4, 10)
dat2 <- data.frame(x, db2, "2000")
names(dat2) <- c("x","db", "df")
dat3 <- data.frame(x, db3, "25")
names(dat3) <- c("x","db", "df")
dat4 <- data.frame(x, db4, "10")
names(dat4) <- c("x","db", "df")
dats <- rbind(dat2, dat3)
dats <- rbind(dats, dat4)
dats$df <- as.factor(dats$df)
levels(dats)
ggplot(data = dats, aes(x = x, y = db, color = df)) +
  geom_line() +
  theme_minimal() +
  ylim(0,.8) + xlim(0, 5) +
  xlab("F statistic") + ylab("Probability") + 
  guides(color = guide_legend("Denominator\nd.f.")) +
  scale_color_manual(values = c("dodgerblue2","#88398a", "deeppink"))

```

Leslie reviews the plot and notices that, while different, the lines do not seem to be **very** different. Chelsea notes that while it is true that there isn't much difference between the lines, the area under the curves is what matters for the p-value cutoff. When the line is just slightly closer to the x-axis, this adds up quickly for the area under the curve. The thresholds for statistical significance (p < .05) for these three lines are `r qf(.95, 4, 2000)` for the 2000 degrees of freedom, `r qf(.95, 4, 25)` for the 25 degrees of freedom, and `r qf(.95, 4, 10)` for the 10 degrees of freedom. Chelsea reminds Leslie that the numerator degrees of freedom has a bigger impact on the significance threshold. For example, an ANOVA with 2 degrees of freedom and the same 2000 degrees of freedom in the denominator would have a threshold for significance of `r qf(.95, 2, 2000)` instead of `r qf(.95, 4, 2000)` for the 4 and 2000 threshold.

### The Kruskal-Wallis test for failing the normality assumption

The Kruskal-Wallis test is used to compare three or more groups when the normal distribution assumption fails for ANOVA. Like several of the tests used when the the outcome is not normally distributed for a t-test, the Kruskal-Wallis test compares ranks among groups. Specifically, the observations are put in order by size and each is assigned a rank. The mean rank for each group is then computed and used to calculate the KW test statistic:

$$
\begin{equation}
KW=\frac{12}{n(n+1)}\sum{n_i(R_i-\frac{n+1}{2})^2}
 (\#eq:cov)
\end{equation}
$$

In this formula, n is the overall sample size, n_i is the sample size for group i and R_i is the mean rank for group i. The null and alternate hypotheses would be:

H0: The mean rank of technology use is the same across educational degree groups.

HA: The mean rank of technology use is not the same across educational degree groups.

```{r}
# compare usetech by degree
kw.usetech.by.degree <- kruskal.test(formula = USETECH ~ DEGREE,
                                     data = gss.2018)
kw.usetech.by.degree
```

The conclusion is that there is a difference in the mean rank for technology use by degree group (KW(`r kw.usetech.by.degree$parameter`) = `r kw.usetech.by.degree$parameter`; p = `r format(kw.usetech.by.degree$p.value, scientific = FALSE)`). Like the ANOVA results, the KW  identifies if there is a difference somewhere among the means but does not identify which groups are different from one another. A post-hoc test like _Bonferroni_ or _Tukey_ could help. For KW, _Dunn's test_ of multiple comparisons is useful for identifying which groups are _statistically significantly_ different from which other groups. 

Dunn's test command `dunn.test()` is available in the dunn.test package. After installing the package, the command requires a method be selected for adjusting the p-value to account for the multiple comparisons. Chelsea suggests Bonferroni since that is one of the methods commonly used with ANOVA:

```{r}
# post-hoc test for usetech by degree
library(dunn.test)
dunn.usetech.by.degree <- dunn.test(gss.2018$USETECH, gss.2018$DEGREE,
                                     method = "bonferroni")
```

Note that the Dunn's test is a rank-sum test just like the Mann-Whitney U and can be interpreted in the same way. In this case it appears that there is _no difference_ in technology use for _graduate vs. bachelor_, _junior college vs. bachelor_, or _junior college vs. graduate_. All other pairings have statistically significant differences. The table shows a z-statistic for each pair computed from the sum of the ranks for the pair. Below the z-statistic is a p-value associated with the z-statistic. The p-value is adjusted using a Bonferroni adjustment, which means it was multiplied by the number of comparisons. In this case, the number of comparisons was 5(5-1)/2 = 10.

Leslie thinks it would be good to visualize the differences in order to better understand the results. She creates a rank variable and plots it:

```{r fig.cap="Ranks of technology use by degree groups."}
# create rank variable for systolic blood pressure
gss.2018$usetech.rank <- rank(gss.2018$usetech)

ggplot(data = subset(gss.2018, !is.na(usetech.rank)), 
       aes(y = usetech.rank, x = DEGREE)) + 
  geom_boxplot(aes(fill = DEGREE), alpha = .8) + 
  geom_jitter(aes(color = DEGREE), alpha = .3) +
  scale_fill_manual(values = c("gray40", "dodgerblue2","#88398a", 
                               "deeppink", "gray80"), guide = FALSE) +
  scale_color_manual(values = c("gray40", "dodgerblue2","#88398a", 
                                "deeppink", "gray80"), guide = FALSE) +
  theme_minimal() + xlab("Degree")

```

Leslie is very confused. She knows that technology use was lower among those in the < high school group, but this plot shows the group as higher ranked. Chelsea explains that the `rank()` command ranks from highest to lowest, so those with 100% technology use are ranked as number 1. With the y-axis numbered as usual from lowest to highest, the graph shows the higher ranks at the bottom and the lower ranks at the top. 

Once this was clarified, Leslie notices that the plot clearly demonstrates the significant differences seen in the post-hoc tests. The three college groups are very similar to one another, and there are differences among the other groups.

### Unlocking achievement 5: Check your understanding

Next to each test, put the ANOVA assumption violation it addresses.

* Brown-Forsyth
* Welch
* Kruskal-Wallis

## Achievement 6: Two-way ANOVA 

One-way ANOVA is useful when there is a single categorical variable (with 3+ categories) and a continuous variable being compared across the categories. What happens when there there are two categorical variables that may both be useful in explaining a continuous outcome? For example, Leslie has read that technology use varies by sex and age. Bobbi says that both of those predictors are available in the data set and they can use ANOVA methods to answer the question. Specifically, two-way ANOVA is useful for situations where means are compared across the categories of two variables. ANCOVA (analysis of covariance) is useful for comparing means across groups while accounting for, or controlling for, some other continuous variable.

### EDA for two-way ANOVA

The boxplots for technology use by degree showed an increase in the percent of time using technology for those with higher educational attainment. Leslie suggests looking at the use of technology by sex:

```{r fig.cap="Distribution of technology use by sex"}
# graph usetech
ggplot(data = subset(gss.2018, !is.na(USETECH)), 
       aes(y = USETECH, x = SEX)) + 
  geom_jitter(aes(color = SEX), alpha = .3) +
  geom_boxplot(aes(fill = SEX), alpha = .8) + 
  scale_fill_manual(values = c("dodgerblue2","deeppink")) +
  scale_color_manual(values = c("dodgerblue2","deeppink")) +
  theme_minimal() + xlab("Sex") + 
  ylab("Percent of time spent using technology")
```

It appears that sex does have some relationship to time spent on technology use. Chelsea explains that two-way ANOVA can be used to determine if degree and sex both have relationships with technology use by themselves and whether they **interact** to explain technology use. That is, does technology use differ by degree differently for males and females. She suggests examining a boxplot with both categorical variables to help make a prediction:

```{r fig.cap="Distribution of technology use by degree and sex"}
# graph usetech
ggplot(data = subset(gss.2018, !is.na(USETECH)), 
       aes(y = USETECH, x = DEGREE)) + 
  geom_boxplot(aes(fill = SEX), alpha = .8) + 
  scale_fill_manual(values = c("dodgerblue2","deeppink")) +
  theme_minimal() + xlab("Degree") + 
  ylab("Percent of time spent using technology")
```

Leslie notices that there appears to be a different pattern of technology use for males and females. Females with less than a high school degree are using technology a lower percent of the time, but then jump past the males in the high school group and stay higher than males for the junior college group as well. Males and females seem to have relatively equal time spent with technology once a bachelor or graduate degree is earned. 

Chelsea explains that this pattern of differences is consistent with an interaction. She suggests that they look at a traditional **means plot** to visualize the idea of an interaction. Although line graphs are ideally not used to display means by groups, they do aid in understanding what an interaction looks like:

```{r fig.cap = "Means plot of technology use by degree and sex."}
ggplot(data = subset(gss.2018, !is.na(USETECH)), 
       aes(y = USETECH, x = DEGREE, color = SEX)) + 
  stat_summary(fun.y = mean, geom="point", size = 2) +
  stat_summary(fun.y = mean, geom="line", aes(group = SEX), size = 1) +
  scale_color_manual(values = c("dodgerblue2","deeppink")) +
  theme_minimal() + xlab("Degree") + ylim(0,100) +
  ylab("Percent of time spent using technology")
```

When the lines in means plots like this one are parallel, this indicates that there is no interaction between the two categorical variables. The mean of the continuous variable is consistently higher or lower for certain groups compared to others. When the means plot shows lines that cross or diverge, this indicates that there is an interaction between the categorical variables. The mean of the continuous variable is different at different levels of the first continuous variable depending on the value of the other continuous variable. For example, mean technology use is lower for females compared to males for the lowest and highest degree categories, but female technology use is higher than male tech use for the three other categories of degree.

Given this interaction and the differences seen in technology use by degree and by sex, Chelsea suggests that the two-way ANOVA will likely find significant relationships for degree, sex, and the interaction between the two. Leslie thinks this sounds logical and starts the NHST process to test it. Leslie suggests a quick look at the values of the means before estimating the ANOVA:

```{r}
# means by degree and sex
use.stats.2 <- gss.2018 %>% 
  group_by(DEGREE, SEX) %>% 
  summarise(m.techuse = mean(USETECH, na.rm = T),
            sd.techuse = sd(USETECH, na.rm = T))
use.stats.2
```

### Two-way ANOVA NHST

#### NHST Step 1: Write the hypotheses 

H0: The mean level of technology use is the same across groups by degree, sex, and their interaction. 

HA: The mean level of technology use is not the same across groups.

#### NHST Step 2: Compute the test statistic

Bobbi steps in to help Leslie at this point. The interaction term is included in the ANOVA command `aov()` by multiplying the two categorical variables. Leslie starts to write `formula = USETECH ~ DEGREE + SEX + DEGREE*SEX` in the `aov()` command, but Bobbi explains that the individual terms are not needed if there is an interaction. The `aov()` command will include individual terms for any variables included in an interaction term. 

Leslie writes the code to get the test statistic:

```{r}
# two-way ANOVA technology use by degree and sex
techuse.by.deg.sex <- aov(formula = USETECH ~ DEGREE * SEX, data = gss.2018)
summary(techuse.by.deg.sex)
```

Chelsea explains that the statistic column in the output is a column of F-statistics. Leslie notices that there are three F-statistics for this ANOVA. One for each of the two individual variables, also called *main effects* and one for the interaction term. 

#### NHST Step 3: Compute the probability for the test statistic (p-value) 

```{r echo = FALSE}
library(broom)
```


The p-values in this case were `r tidy(techuse.by.deg.sex)$p.value[1]`, `r tidy(techuse.by.deg.sex)$p.value[2]`, and `r tidy(techuse.by.deg.sex)$p.value[3]`. These are very tiny p-values, for example, the value of an F-statistic being as large or larger than the F-statistic for the DEGREE main effect happens about `r 100*tidy(techuse.by.deg)$p.value[1]`% of the time when the null hypothesis is true. 

#### NHST Steps 4 & 5: Interpret the probability and write a conclusion 

The mean time spent on technology use was significantly different across degree groups [F(`r tidy(techuse.by.deg.sex)$df[1]`,`r tidy(techuse.by.deg.sex)$df[4]`) = `r tidy(techuse.by.deg.sex)$statistic[1]`; p = `r tidy(techuse.by.deg.sex)$p.value[1]`] indicating these groups likely came from populations with different mean time spent on technology use. Means were also statistically significant by sex [F(`r tidy(techuse.by.deg.sex)$df[2]`,`r tidy(techuse.by.deg.sex)$df[4]`) = `r tidy(techuse.by.deg.sex)$statistic[2]`; p = `r tidy(techuse.by.deg.sex)$p.value[2]`] and there was a statistically significant interaction between degree and sex on technology use [F(`r tidy(techuse.by.deg.sex)$df[3]`,`r tidy(techuse.by.deg.sex)$df[4]`) = `r tidy(techuse.by.deg.sex)$statistic[3]`; p = `r tidy(techuse.by.deg.sex)$p.value[3]`].  The highest mean was `r use.stats.2[9,3]`% of time used for technology for males with graduate degrees. The lowest mean was `r use.stats.2[2,3]`% of the time for females with less than a high school diploma. The interaction between degree and sex shows that time spent on technology use increases more quickly for females with both males and females eventually having high tech use in the top two education groups. 

### Post-hoc test for two-way ANOVA

After a little searching, Leslie is disappointed to find that the Bonferroni post-hoc test is not available in R for two-way ANOVA, but she does find that the Tukey Honestly Significant Difference test still works. She writes a single line of code to determine which groups have statistically significantly different mean technology use:

```{r}
# Tukey post-hoc test
TukeyHSD(techuse.by.deg.sex)
```

Leslie finds that there are so many groups with significant differences that it seems more useful to just include the box plot from the exploratory analysis or the means plot, a table of p-values for the comparisons, and mention any interesting overall patterns in the comparisons. Bobbi and Chelsea agree. Leslie finds that there are significant differences between males and females in the high school and junior college groups, but that males and females are not significantly different across the other educational groups. She also finds that the college groups spent significantly more time using technology than the two other groups, but were not statistically significantly different from each other. Overall it appears that higher education groups spend more time using technology for males and females and that high school and junior college educated females spend more time using technology than males with these same education levels.

### Two-way ANOVA assumptions

Bobbi mentions that the assumptions of homogeneity of variances and normality are also applicable in two-way ANOVA. She explains that normality would be a little trickier to test by looking at each group since there are five degree groups, two sex groups, and 10 degree-by-sex groups (e.g., male and < high school). Instead of checking normality one group at a time, when there are a large number of groups in an ANOVA model, this assumption can be checked by examinging the **residuals**. The residuals are the distances between the value of the outcome for each person and the value of the group mean for that person. When the residuals are normally distributed, this indicates that the values in each group are normally distributed around the group mean. 

#### Testing the normality assumption

Leslie looks at the `techuse.by.deg.sex` object in the environment pane sees the residuals. She uses a Shapiro-Wilk test to check normality statistically and then plots the residuals for a visual assessment:

```{r}
# statistical test of normality for groups
shapiro.test(techuse.by.deg.sex$residuals)
```

Leslie reminds herself that the null hypthesis for the Shapiro-Wilk test is that the distribution is normal. By rejecting this null hypothesis with a tiny p-value, the assumption is failed. Leslie graphs the residuals to confirm. She finds that the `ggplot()` command does not work with the ANOVA :

```{r fig.cap = "Distribution of residuals from tech use by degree and sex ANOVA."}
# plotting the residuals
ggplot(data = data.frame(techuse.by.deg.sex$residuals), aes(x = techuse.by.deg.sex.residuals)) + 
  geom_histogram(fill = "#88398a", col = "grey") + 
  theme_minimal() + 
  labs(x = "Residuals") 

```

The residuals do not appear to be normally distributed. Instead, Leslie notices, they are bimodal, with some large residuals at the lower end and some at the upper end of the range. This indicates that some observations are quite far below or above the group mean value for their group. 

#### Testing the homogeneity of variances assumption

Leslie remembers the `leveneTest()` command from the car package can be used to test the null hypothesis that the variances are equal. Because she does not need the car package open for anything else, she uses `car::` to open the package just for the `leveneTest()` command to test this assumption:

```{r}
# Levene test for ANOVA
car::leveneTest(y = USETECH ~ DEGREE*SEX, center = mean, data = gss.2018)
```

The statistically significant results indicate that the null hypothesis is rejected and therefore the equal variances assumption is not met. 

### Alternatives when two-way ANOVA assumptions fail 

Bobbi recently read a textbook that suggested using a Friedman test [@macfarland2016friedman]. The book described the test as conducting ANOVA on ranked values (like the Mann-Whitney U and Welch one-way ANOVA tests). Chelsea and Leslie think this sounds good and look up the code to conduct the test. In R, the Friedman test can be conducted using `friedman.test()` on a set of means for each group in a summary data format rather than raw data:

```{r}
# Friedman two-way ANOVA for ranked data
# R command requires summary data
agg.gss.2018 <- aggregate(!is.na(gss.2018$USETECH),
                by = list(deg = gss.2018$DEGREE,
                          sex = gss.2018$SEX),
                FUN = mean)

# Friedman test
tech.by.deg.sex.f <- friedman.test(agg.gss.2018$x, agg.gss.2018$deg, agg.gss.2018$sex)
tech.by.deg.sex.f

```

Leslie, Bobbi, and Chelsea are surprised that the Friedman test finds no statistically significant difference in the ranks of technology use by degree and sex [$\chi^2$(`r tech.by.deg.sex.f$parameter`) = `r tech.by.deg.sex.f$statistic`; p = `r tech.by.deg.sex.f$p.value`]. Given the means plots and box plots developed in the exploratory data analysis for two-way ANOVA, this seems like an unusual result. Bobbi does a little digging and finds a number of manuscripts suggesting that the Friedman test is not a great option for addressing failed assumptions because of the way the ranks are computed [@zimmerman1993relative;@harwell1994monte]. 

Instead of using Friedman, another suggested method is to compute the ranks of the outcome and conduct the two-way ANOVA on the ranked outcome variable. Luckily, Leslie computed ranks already for the Dunn's test earlier and tries out the two-way ANOVA code with the transformed outcome variable:

```{r}
# two-way ANOVA technology use by degree and sex
techuse.by.deg.sex.t <- aov(formula = usetech.rank ~ DEGREE * SEX, data = gss.2018)
summary(techuse.by.deg.sex.t)
```

The results are again different from the original two-way ANOVA, but they do make more sense than the Friedman results. Based on an ANOVA of the ranked technology use variable, degree has a significant relationship with technology use but sex does not and the interaction of degree and sex do not. Plots of the ranks might help determine what is going on:


```{r}
# boxplot of usetech ranks
ranks.box <- ggplot(data = subset(gss.2018, !is.na(USETECH)), 
       aes(y = usetech.rank, x = DEGREE)) + 
  geom_boxplot(aes(fill = SEX), alpha = .8) + 
  scale_fill_manual(values = c("dodgerblue2","deeppink")) +
  theme_minimal() + xlab("") + ylim(0,1100) + 
  ylab("Ranked percent of time\nspent using technology") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

# means plot of usetech ranks 
ranks.means <- ggplot(data = subset(gss.2018, !is.na(USETECH)), 
       aes(y = usetech.rank, x = DEGREE, color = SEX)) + 
  stat_summary(fun.y = mean, geom="point", size = 2) +
  stat_summary(fun.y = mean, geom="line", aes(group = SEX), size = 1) +
  scale_color_manual(values = c("dodgerblue2","deeppink")) +
  theme_minimal() + xlab("Degree") + ylim(0,1100) +
  ylab("Ranked percent of time\nspent using technology")

# both plots together
gridExtra::grid.arrange(ranks.box, ranks.means)

```

The plots show a bigger difference from one degree group to the other than between males and females in any given group, so the significant degree variable makes sense. There still appears to be some difference by sex, but it does not rise to the level of statistical significance in this case. 

Another strategy sometimes used when two-way ANOVA assumptions are not met is a Box-Cox transformation on the outcome variable. The Box-Cox power transformations were developed to reduce the non-normality of residuals, so they may be useful here. The original paper by Box and Cox does a nice job of explaining the transformations [@box1964analysis] and there are numerous tutorials on the use of these transformations.

## Chapter summary

### Achievements unlocked in this chapter: Recap

After reading this chapter and following along, Leslie (and you) has learned and practiced: 

#### Achievement 1 recap: Using graphics and descriptive statistics to make a prediction

Prior to conducting inferential statistical tests like chi-squared, it is useful to get some idea of the characteristics and relationships in your data. Descriptive statistics and graphs, or exploratory data analysis (EDA), can serve two purposes: (1) understand the people, things, or phenomena you are studying better, and (2) make an educated prediction about the likely results of a statistical test, which can help identify issues if (or when) the test is not properly conducted.

#### Achievement 2 recap: One-way analysis of variance (ANOVA)

One-way ANOVA compares the means of three or more groups. Significant results indicate that the differences among the means likely reflects a difference in means from the populations the samples came from. One-way ANOVA is an omnibus test (like chi-squared), identifying significant relationships but providing no measure of the strength of the association between variables. 

#### Achievement 3 recap: Post-hoc testing and planned comparisons

Following a significant ANOVA result, planned comparisons or a post-hoc test like Bonferroni or Tukey HSD can identify significant differences between pairs of means. Planned comparisons are typically used when there are hypothesized differences among the groups while post-hoc tests are used when there are no specific hypotheses about differences among group means. 

#### Achievement 4 recap: ANOVA assumptions

Statistical tests rely on underlying assumptions about the characteristics of the data. When these assumptions are not met, the results may not reflect the true relationships among the variables. For ANOVA, four assumptions need to be met: (1) continuous variable and three or more independent groups, (2) independent observations, (3) normal distribution in each group, and (4)  equal variances for each group. 

#### Achievement 5 recap: Alternatives when ANOVA assumptions are not met

When assumptions are not met for ANOVA, there are alternative tests that account for the data not meeting assumptions. When the homogeneity of variances assumption is not met, two alternative F-statistics are good options: Brown-Forsythe and Welch. When the normality assumption is not met, the Kruskal-Wallis test is an alternative test that compares the mean ranks across groups.

#### Achievement 6 recap: Two-way ANOVA

Two-way ANOVA is useful for comparing means across groups of two categorical variables separately and as an interaction. This type of statistical model can answer the question of how tech use varies by education and by sex and by the interaction of education and sex. Assumption checking and post-hoc tests apply in two-way ANOVA in similar ways as they did in one-way ANOVA.

### Chapter exercises 

The coder and hacker exercises are an opportunity to apply the skills from this chapter to a new scenario or a new data set. The coder edition will evaluate your application of the commands learned in this chapter (and earlier chapters) to similar scenarios to those in the chapter; the hacker edition will evaluate your use of the procedures from this chapter in new scenarios, usually going a step beyond what was explicitly explained. 

Before picking the coder or hacker version, check your knowledge. We recommend the coder edition if you answer all 5 multiple choice questions correctly by your third try and the hacker edition if you answer at least 3 of the 5 multiple choice questions correctly on your first try the rest correctly on your first or second try.

Q1: What is the primary purpose of ANOVA?

a. Comparing means across 3 or more groups* 
b. Comparing medians across 3 or more groups 
c. Examining the relationship between two categorical variables 
d. Identifying normally distributed data 

Q2: Which of the following assumptions does NOT apply to ANOVA? 

a. Independent observations 
b. Normal distribution of continuous variable  
c. Homogeneity of variances 
d. Includes one bivariate variable*  

Q3: How many pairwise comparisons would there be for an ANOVA with four groups? 

a. 16 
b. 4 
c. 12 
d. 6* 

Q4: Apply a Bonferroni adjustment to a p-value of .01 if the analyses included 6 pairwise comparisons. If the threshold for statistical significance were .05, would the adjusted p-value be significant? 

a. Yes 
b. No* 

Q5: In which situation would you use planned comparisons?

a. After a significant ANOVA to compare each pair of means 
b. Instead of an ANOVA when the data did not meet the normality assumption 
c. When you have to choose between two categorical variables 
d. When you conduct an ANOVA and have hypotheses about which sets of means are different from one another

#### Chapter exercises: Coder edition 

Depending on your score in the knowledge check, choose either the coder or hacker edition of the chapter exercises. Use the data from this chapter and the appropriate tests to examine technology use by happiness. 

1) Open the data using the strategy shown in this chapter 
2) Clean the happiness, sex, and tech use variables so they have clear variable names, category labels, and missing value coding 
3) (**A1**) Use graphics and descriptive statistics to examine tech use on its own, by sex, and by happiness 
4) (**A1**) Based on the graphs and statistics from question 3, make a prediction about what you would find when you compare tech use across groups for happiness and for sex 
5) (**A2**, **A5**) Conduct the appropriate test to compare mean tech use across happiness groups. If the F-statistic is significant, use a post-hoc test to determine which means are statistically significantly different from each other. Interpret your results. 
6) (**A3**, **A5**) Check assumptions for the ANOVA and conduct an appropriate alternate analysis if it does not pass assumptions. Interpret your results. 
7) (**A6**) Conduct a two-way ANOVA using sex and happiness. Interpret results and check assumptions.

#### Chapter exercises: Hacker edition

Complete the coder edition and:

* Use the graphics in #3 to hypothesize which happiness groups have different tech use means. Write a appropriate planned comparisons instead to test your hypotheses and conduct an appropriate planned comparison analysis instead of post-hoc tests in #5. 
* Conduct and interpret post-hoc test for the two-way ANOVA developed in #7 

### BOX(ES)

#### (BOX) Leslie's stats stuff: Type I and Type II errors

<img align = "left" src = "avatars\leslie.gif" style="PADDING-RIGHT: 30px">

There are four possible outcomes to a statistical test: 

```{r echo = FALSE, message = FALSE}
library(descr)
row.col.vals <- matrix(c('True', 'Type I error', 
                          'Type II error', 'True'), 
                        nrow = 2, dimnames = list(c("Fail to reject", "Reject"),
                                           c("True",
                                                        "False")))

library(htmlTable)
htmlTable(row.col.vals,
          header =  paste(c("True", "False"), ""),
          rnames = paste(c("Failed to reject", "Rejected"), ""),
          cgroup = "In reality, the null hypothesis is...",
          n.rgroup = 2,
          rgroup = "Hypothesis test")


```

A Type II error, also called $\beta$, is when there is a relationship but the study did not detect it. A Type I error, also called $\alpha$, is when there is **no relationship** but the study detects one. The $\alpha$ for Type I error is also the threshold set for statistical significance. The threshold for statistical significance is the amount of uncertainty tolerated in concluding whether or not a result is _statistically significant_. If $\alpha$ is .05, this is a willingness to risk a 5% chance of making a Type I error of rejecting the null hypothesis when it should be retained. Increasing sample size and decreasing the threshold for _statistical significance_ ($\alpha$) can aid in decreasing Type I and II errors. 



> Gamification idea for Chapter 7: Performance on student work in the achievements unlocked, are you a coder or hacker, and chapter exercises in this chapter could be included as part of earning a continuous data analysis badge. Combine scores from this chapter and chapters 10 and 11. Set thresholds for earning a bronze, silver, or gold badge. Integrate badge earnings into Blackboard or other platforms students use to share their portfolios and achievements with prospective employers or degree programs. Create a class contest for students to create a decision tree for choosing descriptive and bivariate statistics depending on variable type and assumptions. Students can work as individuals or groups. Decision trees are posted online and the class votes for the "best" decision tree. The individual or group responsible gets additional points for the chapter 6 exercises. 


